Project Diary

### 2014-06-11 thesis/cross_correlation_gaussian.tex
Move matlab files to a separate folder.

\chapter{Correlation Matrix of Gaussian Returns}\label{chp:Gaussian}
In this chapter we present some results about the cross-correlation
matrix of Gaussian returns. This assumption is of course an
over-simplification, but nevertheless lends some insight into the
problem.

Our primary interest is in the influence of autocorrelations on the
cross-correlation matrix. \S\ref{sec:GCC-analytical} and
\S\ref{sec:GCC-numerical} discuss the distribution of the matrix
elements and the eigenvalues, respectively.

\section{Distribution of the Matrix Elements}\label{sec:GCC-analytical}
If the returns follow a zero-mean Gaussian distribution, 
i.e. $\vec{r}_t \sim N(0, \Sigma)$, and are not
auto-correlated, then $RR'$ is a Wishart matrix with
the following distribution law \cite{Anderson2003}:
\begin{equation}
  \label{eq:WishartPDF}
  f_{RR'}(X) = { |\det X|^{(T-N-1)/2} \exp\left(-{1 \over 2}\tr
      \Sigma^{-1}X \right)
    \over
    2^{NT/2}\pi^{N(N-1)/4}|\det \Sigma|^{T/2}
    \prod_{i=1}^N \Gamma\left[(n+1-i)/2\right]
  }
\end{equation}
where $\Sigma$ is the true covariance matrix of the returns.

When $\vec{r}_t$ are indeed auto-correlated, $RR'$ no longer has the
Wishart distribution. The joint distribution function of its elements
can be expressed in terms of the Wishart PDF and the
auto-correlations. We show this in appendix
\ref{app:pdf_gaussian1}. In the rest of this section, we derive an
approximate expression for the asymptotic distribution of these matrix
elements.

In the following we denote the diagonal elements as $C_{ii}$ and the
non-diagonal elements as $C_{ij}$ ($i \neq j$). We assume that
$\var(a_{i,t}) = \sigma^2$ for any $i$ and $t$, and $a_{i,t}$ are not
autocorrelated, i.e. $\text{corr}(a_{i,t}, a_{i, t'}) = 0$ for any
$i$, $t$ and $t'$. Then we can write
\begin{eqnarray*}
  r_{i,t} r_{j, t} &=& \left[
    \phi_1 r_{i, t-1} + a_{i, t}
  \right] \left[
    \phi_1 r_{j, t-1} + a_{j, t}
  \right] \\
  C_{ij} &=& {1 \over T}\sum_{t=1}^T r_{i,t} r_{j, t} \\
  &=& \phi_1^2 {1 \over T}\sum_{t=1}^Tr_{i,t-1} r_{j, t-1} +
  \phi_1{1 \over T}\sum_{t=1}^T\left(
    r_{i, t-1} a_{j, t} + r_{j, t-1} a_{i, t}
  \right) + {1 \over T}\sum_{t=1}^T a_{i,t} a_{j,t}
\end{eqnarray*}
We note that the two sums $\sum_{t=1}^T r_{i,t} r_{j, t}$ and
$\sum_{t=1}^T r_{i,t-1} r_{j, t-1}$ only differ by the first and the
last addend, which is negligible for sufficiently large T. Thus we
have
\begin{equation}\label{eq:Offdiag1}
  \begin{aligned}
    (1 - \phi_1^2)C_{ij} &\approx& \phi_1 {1 \over T} \sum_{t=1}^T
    \left(r_{i, t-1} a_{j, t} + r_{j, t-1} a_{i, t} \right)
    + {1 \over T}\sum_{t=1}^T a_{i,t} a_{j,t}
  \end{aligned}
\end{equation}
Now we write the AR(1) process $r_{i,t}$ as an infinite moving-average
process:
\begin{eqnarray*}
  r_{i, t} &=& \phi_1 r_{i, t-1} + a_{i,t} \\
  (1 - \phi_1 B) r_{i, t} &=& a_{i,t} \\
\end{eqnarray*}
where $B$ is the back-shift operator. Then it follows from the above
equation
\begin{eqnarray*}
  r_{i,t} &=& {1 \over 1 - \phi_1 B} a_{i,t} \\
  &=& \sum_{k=0}^\infty \phi_1^k B^k a_{i,t} \\
  &=& \sum_{k=0}^\infty \phi_1^k a_{i,t-k} \\
\end{eqnarray*}
where it is left implicit that $a_{i, t}$ with $t \leq 0$ is zero (in
words, this implies that the $r_{i,t}$ process is not affected at all
by events before $t = 1$).

Substituting this into eq.\ref{eq:Offdiag1} for $r_{i,t-1}$ yields
\begin{equation}
  \label{eq:Offdiag2}
  \begin{aligned}
  (1 - \phi_1^2)C_{ij} &\approx
  {1 \over T} \sum_{t=1}^T \sum_{k=0}^\infty\phi_1^{k+1} (a_{i, t-k-1}
  a_{j, t} + a_{j, t-k-1} a_{i, t})
  + {1 \over T} \sum_{t=1}^T a_{i,t} a_{j,t} \\
  &=
  {1 \over T} \sum_{t=1}^{T} \sum_{k=1}^{t-1}\phi_1^k (a_{i, t-k}
  a_{j, t} + a_{j, t-k} a_{i, t})
  + {1 \over T} \sum_{t=1}^{T} a_{i,t} a_{j,t}\\
  \end{aligned}
\end{equation}

Given two Gaussian random variables $x$ and $y$ with zero mean and
covariance matrix 
\begin{equation*}
  \Sigma =
  \begin{pmatrix}
    1 & \rho \\
    \rho & 1 \\
  \end{pmatrix}
\end{equation*}

The PDF of $xy$ can be found by considering $P(xy < z)$:
\begin{eqnarray*}
  P(xy < z) &=& \left(\int_0^\infty dx \int_0^{z/x} dy
    +\int_{-\infty}^0 dx \int_{z/x}^\infty dy \right)
    {1 \over 2\pi \sqrt{1 - \rho^2}} \\
    &&
    \exp\left[
      -{x^2 -2\rho xy + y^2
        \over
        2\sigma^2 (1 - \rho^2)} 
    \right] \\
  f(z; \sigma, \rho) &=& {d \over dz} P(xy < z)\\
  &=& {1 \over \pi \sqrt{1 - \rho^2}} \exp\left[
    {\rho z \over \sigma^2 (1 - \rho^2)}\right] K_0\left[
    {|z| \over \sigma^2 (1 - \rho^2)}
  \right]
\end{eqnarray*}
where $K_n(z')$ is the modified Bessel function of the second
kind. It is worth taking note that, when $\rho \neq 0$,
$f(z; \sigma, \rho)$ is not symmetric with respect to
$z$. As a result, if $\rho > 0$, the mean of $f(z;
\sigma, \rho)$ is positive, and vice versa.

Secondly, because $|\rho| < 1$ and
\begin{equation*}
  K_0(z) \sim \sqrt{\pi \over 2z} e^{-z}
\end{equation*}
as $z \to \infty$ \cite{Olver:2010:NHMF}, all the moments of
$f(z; \sigma, \rho)$ are finite, implying the
applicability of the Lyapunov central limit theorem provided that $T$
is large, which is very often the case and what we assume here.

With this in mind, we observe that $\phi_1$ only affects the first sum in
\ref{eq:Offdiag2}. If $a_{i,t-k}$ and $a_{j,t}$ are not correlated for
non-zero $k$, $\phi_1$ will not affect the mean of
$C_{ij}$. Furthermore, it is also clear from equation
\ref{eq:Offdiag2} that the variance of $C_{ij}$ is
always increased by a non-zero $\phi_1$, regardless of the sign of
$\phi_1$. We compute this increment in the following.

In light of the above expression for $f(z; \sigma,
\rho)$, we rewrite equation \ref{eq:Offdiag2} as
\begin{equation}\label{eq:Cij_dist}
  \begin{aligned}
    (1 - \phi_1^2)C_{ij} \approx &
    {1 \over T}
    \sum_{t=1}^{T} \sum_{k=1}^{t-1}\phi_1^k (a_{i, t-k} a_{j, t} + a_{j,
      t-k} a_{i, t}) \\
    & + {1 \over T} \sum_{t=1}^{T} \sigma^2 (1 - \rho^2) {a_{i,t} \over
      \sigma \sqrt{1 - \rho^2}} {a_{j,t} \over \sigma \sqrt{1 - \rho^2}} \\
  \end{aligned}
\end{equation}
In addition, we assume
\begin{equation*}
  \text{corr}(a_{i, t-k}, a_{j, t}) = \left\{
    \begin{array}{l l}
      1 & \text{ if $i=j$ and $k = 0$ }\\
      \rho & -1 < \rho < 1. \text{ if $i \neq j$ and $k = 0$ }\\
      0 & \text{otherwise}
    \end{array}
  \right.
\end{equation*}
Then, because $a_{i, t-k}$ and $ a_{j, t}$ with $i \neq j$ and $k
> 0$ are not correlated, the mean of $a_{i, t-k} a_{j, t}$ is 0
($f(z; \sigma, 0)$ is symmetric), and the variance of it
can be found to be $\sigma^6$ using formula (10.43.19) of \cite{NIST:DLMF}:
\begin{equation*}
  \int_0^\infty dt K_\nu(t) t^{\mu-1} = 2^{\mu-2}
  \Gamma\left(
    {\mu + \nu \over 2}
  \right) \Gamma\left(
    {\mu - \nu \over 2}
  \right)
\end{equation*}
On the other hand, $a_{i,t}/\sigma \sqrt{1 - \rho^2}$ and
$a_{j,t}/\sigma \sqrt{1 - \rho^2}$ have variance $1/(1 - \rho^2)$
and are correlated - $\text{corr}(a_{i,t}, a_{j,t}) = \rho$. The mean
of $a_{i,t}a_{j,t}/\sigma^2 (1 - \rho^2)$ can be found using formula
(10.43.22) of \cite{NIST:DLMF}, given that $-1 < \rho < 1$:
\begin{eqnarray*}
  \int_0^\infty t^{\mu - 1} e^{-at} K_\nu(t) dt &=& (\pi/2)^{1/2}
  \Gamma(\mu + \nu) \Gamma(\mu - \nu)(1 - a^2)^{-\mu/2 + 1/4} \times\\
  && P^{-\mu+1/2}_{\nu-1/2} (a)
\end{eqnarray*}
where $P^\mu_\nu(\cdot)$ is Ferrers function of the first
kind\footnote{ Ferrers function of the first kind is defined through
  the hypergeometric functon $F(a, b; c; z)$ \cite{NIST:DLMF}:
  \begin{equation}
    \label{eq:Ferrers_1st}
    \mathop{\mathsf{P}^{\mu}_{\nu}\/}\nolimits\!\left(x\right)=\left(\frac{1+x}{1-
        x}\right)^{\mu/2}\mathop{\mathbf{F}\/}\nolimits\!\left(\nu+1,-\nu;1-\mu;\tfrac
      {1}{2}-\tfrac{1}{2}x\right).
  \end{equation}
}. The result is
\begin{eqnarray*}
  \E\left[{a_{i,t}a_{j,t} \over \sigma^2 (1 - \rho^2)}\right]
  &=&
  {1 \over \sqrt{2\pi} (1 - \rho^2)^{5/4}} \left[
    P^{-3/2}_{-1/2}(-\rho) - P^{-3/2}_{-1/2}(\rho)
  \right]
\end{eqnarray*}
Similarly, the variance of $a_{i,t}a_{j,t}/\sigma^2 (1 - \rho^2)$ is
found to be
\begin{eqnarray*}
  v^2(\rho) &=&
  {4 \over \sqrt{2\pi} (1 - \rho^2)^{7/4}} \left[
    P^{-5/2}_{-1/2}(\rho) + P^{-5/2}_{-1/2}(-\rho)
  \right] \\
  && - \E^2\left[{a_{i,t}a_{j,t} \over
      \sigma^2 (1 - \rho^2)}\right]
\end{eqnarray*}
Now we can apply the Lyapunov central limit theorem
\cite{Billingsley1995} to the sum in equation \ref{eq:Cij_dist}
and write down the asymptotic Gaussian distribution of $C_{ij}$:
\begin{equation*}
C_{ij} \sim N(\mu'_X, \sigma'_X)
\end{equation*}
where
\begin{eqnarray}
  \mu'_X &=& {\sigma^2 \over \sqrt{2\pi} (1 - \phi_1^2)(1 -
    \rho^2)^{1/4}} \left[ P^{-3/2}_{-1/2}(-\rho) -
    P^{-3/2}_{-1/2}(\rho)
  \right] \label{eq:gaussian_mean}\\
  \sigma'^2_X &=& {1 \over (1 - \phi_1^2)^2}\left[
    \sum_{t=1}^T \sum_{k=1}^{t-1} 2\left(
      \phi_1^k \over T
    \right)^2 \sigma^6 + \sum_{t=1}^T
    {\sigma^4 (1 - \rho^2)^2 \over T^2} v^2(\rho)
  \right] \nonumber \\
  &=& {2 \sigma^6 \over T (1 - \phi_1^2)^2} \left[
    {\phi_1^2 \over 1 - \phi_1^2} -
    {\phi_1^2 (1 - \phi_1^{2T}) \over
      T(1 - \phi_1^2)}
  \right] + {\sigma^4 (1 - \rho^2)^2 v^2(\rho) \over
    T (1 - \phi_1^2)^2} \nonumber \\
  &\approx& {2 \sigma^6 \phi_1^2 \over T (1 - \phi_1^2)^3}
  + {\sigma^4 (1 - \rho^2)^2 v^2(\rho) \over
    T (1 - \phi_1^2)^2} \label{eq:gaussian_variance}
\end{eqnarray}
Equation \ref{eq:gaussian_mean} tells that, if two return series $i$
and $j$ are not correlated, auto-correlation in the returns does not
introduce a bias into the estimation of the cross-correlation; if,
however, the return series are indeed correlated, auto-correlation
in the returns rescales the cross-correlation through a multiplicative
factor $1/(1 - \phi_1^2)$.

In addition, equation \ref{eq:gaussian_variance} tells that
auto-correlation in the returns always makes the cross-correlation
estimation more noisy. Auto-correlation not only rescales the variance
of the no-autocorrelation estimation by $1/(1 - \phi_1^2)$ but even
adds an extra term ${2 \sigma^6 \phi_1^2 \over T (1 - \phi_1^2)^3}$.

% What we have obtained is a quantitative description of what has been
% said before: The mean of the non-diagonal entries of the
% cross-correlation matrix is not affected by autocorrelations in the
% returns, given that innovations $a_{i,t}$ of different assets and
% different times are un-correlated; the variance, however, is always
% increased by autocorrelations no matter the autocorrelation is
% positive or nagative. The increment is in the form of an additive term
% that scales as $\phi_1^2 / (1 - \phi_1^2)^3$.

We may apply a similar treatment to the diagonal elements of the
covariance matrix, which we denote as $C_{ii}$ here:
\begin{eqnarray*}
  C_{ii} &=& {1 \over T} \sum_{t=1}^T r^2_{it} \\
  &=& {1 \over T} \sum_{t=1}^T \sum_{l=0}^{t-1}\phi_1^l a_{i, t-l}
  \sum_{k=0}^{t-1}\phi_1^k a_{i, t-k} \\
  &=& {1 \over T} \sum_{t=1}^T \left[
    \sum_{k=0}^{t-1} \phi_1^{2k} a^2_{i, t-k} +
    \sum_{k,l = 0}^{t-1} \phi_1^{k+l} a_{i, t-k} a_{i, t-l}
  \right]
\end{eqnarray*}
By the Lyapunov central limit theorem under the assumption of large T,
the asymptotic distribution of $C_{ii}$ is Gaussian, the mean and
variance being
\begin{eqnarray}
  \E(C_{ii}) &=& {1 \over T}\left[
    \sum_{k=0}^{t-1} \phi_1^{2k} \sigma^2
  \right] \nonumber \\
  &=& {\sigma^2 \over (1 - \phi_1^2) T} \left[
    T - {\phi_1^2 (1 - \phi_1^{2T}) \over 1 - \phi_1^2}
  \right] \nonumber \\
  & \approx & {\sigma^2 \over 1 - \phi_1^2} \left[
    1 - {\phi_1^2 \over T}
  \right] \label{eq:gaussian_cii_mean}
\end{eqnarray}
and
\begin{eqnarray}
  \var(C_{ii}) &=& \sum_{t=1}^T \left[
    \sum_{k=0}^{t-1} {\phi_1^{4k} \sigma^4 \over T^2} 2 +
    \sum_{k,l=0}^{t-1} {\phi_1^{2(k+l)} \over T^2} \sigma^6
  \right] \nonumber \\
  &=& \sum_{t=1}^T \left[
    {2 \sigma^4 \over T^2} {1 - \phi_1^{4t} \over 1 - \phi_1^4} +
    {\sigma^6 \over T^2} \left(
      {1 - \phi_1^{2t} \over 1- \phi_1^2}
    \right)^2 \right] \nonumber \\
  &=& {2 \sigma^4 \over T (1 - \phi_1^4)} -
  {2 \sigma^4 \phi_1^4 (1 - \phi_1^{4T}) \over T^2(1 - \phi_1^4)^2} +
  \nonumber \\
  && {\sigma^6 \over T (1 -\phi_1^2)^2} -
  {2 \sigma^6 \phi_1^2 (1 - \phi_1^{2T}) \over T^2 (1 - \phi_1^2)^3} +
  {\sigma^6 \phi_1^4 (1 - \phi_1^{4T}) \over T^2 (1 - \phi_1^2)^2 (1 -
    \phi_1^4)} \nonumber \\
  &\approx& {2 \sigma^4 \over T (1 - \phi_1^4)} + {\sigma^6 \over T (1
    -\phi_1^2)^2} \label{eq:gaussian_cii_variance}
\end{eqnarray}
From equation \ref{eq:gaussian_cii_mean} we see that auto-correlation
in the returns increases the variance of the return series; and from
equation \ref{eq:gaussian_cii_variance} we see that the variance of
that variance estimation is also increased by
auto-correlations. Moreover, we note that $\var(C_{ii})$ scales with T
approximately as $1/T$, similar to the behavior of
$\var(C_{ij})$. This is to be compared with the case of GARCH returns
discussed in chapter \ref{chp:CrossCorrelationFat}.

% As for the diagonal elements of the cross-correlation matrix,
% i.e. $\var(r_{it})$, from the representation of $r_{it}$ as an
% inifinite moving average process, one can deduce immediately
% \begin{equation*}
%   \text{var}(r_{i,t}) \approx {\sigma^2 \over 1 - \phi^2}
% \end{equation*}
% as is mentioned earlier in equation \ref{eq:gaussian_Cii_variance}.

% However, to write $f_A(RMM'R')$ as a function of the eigen
% values of $RMM'R'$ is very involved and no theoretical results are
% known.

% In the more general case
% \begin{eqnarray*}
%   a_{i,t} &=& \left(
%     1 - \sum_{k=1}^p \phi_{i,k} B^k
%   \right) r_{i,t} \\
%   &=& \Phi_i(B)} r_{i,t
% \end{eqnarray*}
% where $B$ denotes the back shift operator acting on subscript $t$, we
% can write
% \begin{eqnarray*}
%   A &=&
%   \begin{pmatrix}
%     \Phi_1(B) & &\\
%     & \ddots & \\
%     && \Phi_N(B)
%   \end{pmatrix} R \\
%   &=& \Phi(B) R
% \end{eqnarray*}

>>>>

\chapter{Covariance Matrix of Gaussian Returns}
\label{chp:Gaussian}
In this chapter we present some results about the covariance
matrix of Gaussian returns. The assumption of Gaussian returns is of
course an over-simplification of reallity, but nevertheless lends some
insight into the problem of elements' and eigenvalues' distributions of
a covariance matrix.
Our primary interest is in the influence that auto-correlations in
returns exert on the covariance matrix. Section \ref{sec:GCC-analytical} and
section \ref{sec:GCC-numerical} discuss how the distributions of the matrix
elements and the eigenvalues are affected by such autocorrelations,
respectively.

\section{Distribution of the Matrix Elements}
\label{sec:GCC-analytical}
In this section we study the distribution of the elements of a
covariance matrix $C = \mtx{RR'}/T$, where
\begin{eqnarray*}
  \mtx R &=&
  \begin{pmatrix}
    r_{11} & r_{12} & \cdots & r_{1T} \\
    r_{21} & r_{22} & \cdots & r_{2T} \\
    \vdots & \vdots & \ddots & \vdots \\
    r_{N1} & r_{N22} & \cdots & r_{NT} \\
  \end{pmatrix}
\end{eqnarray*}
In words, an element $r_{it}$ of $\mtx R$ is the return of asset $i$
at time $t$, with $i=1,\cdots, N$ and $t = 1, \cdots, T$. If each
column of $\mtx R$ follows a zero-mean Gaussian distribution, i.e.
\begin{eqnarray*}
  \begin{pmatrix}
    r_{1t} \\
    \vdots \\
    r_{Nt}
  \end{pmatrix} \sim N(0, \mtx \Sigma)
\end{eqnarray*}
for all $t = 1, \cdots, T$, and none of the return series is
auto-correlated, i.e. $\text{corr}(r_{it}, r_{i,t'}) = 0$ for all $i =
1, \cdots, N$ and $t \neq t'$, then $\mtx{RR'}$ is a Wishart matrix whose
probability density function is well known \cite{Anderson2003}.

When auto-correlations are indeed present in the returns, $\mtx{RR'}$
no longer follows the Wishart distribution. However, the joint distribution
function of its elements can be expressed in terms of the Wishart \gls{pdf}
and the auto-correlations. We show this in appendix
\ref{app:pdf_gaussian1}. Also, in appendix
\ref{chp:gaussian_elements_dist} we derive an approximate expression
for the asymptotic distribution of these matrix elements, assuming
$r_{it}$ is an \gls{ar}(1) process \footnotemark, i.e. $r_{it} =
\phi r_{i, t-1} + a_{it}$, and
\footnotetext{
  It is straight forward to derive the auto-correlation function
  $\varrho_k$ of an \gls{ar}(1) process with autoregressive
  coefficient $\phi$:
  \begin{eqnarray*}
    \varrho_k &=& \text{corr}(r_{it}, r_{i, t-k}) \\
    &=& \phi \varrho_{k-1} \\
    &=& \phi^k
  \end{eqnarray*}
  Let $\tau$ denote the time lag at which $\varrho_{\tau} = 1/2$, then
  it follows $\tau = -\ln 2/\ln \phi$.
}
\begin{eqnarray*}
  \begin{pmatrix}
    a_{1t} \\
    \vdots \\
    a_{Nt}
  \end{pmatrix} \sim N(0, \mtx \Sigma)
\end{eqnarray*}
where
\begin{eqnarray*}
  \mtx \Sigma &=& \sigma^2
  \begin{pmatrix}
    1 & \rho & \cdots & \rho \\
    \rho & 1 & \cdots & \rho \\
    \vdots & \vdots & \ddots & \vdots \\
    \rho & \rho & \cdots & 1
  \end{pmatrix}
\end{eqnarray*}
where $-1 < \rho < 1$ is a constant parameter describing the correlation
between $a_{it}$ and $a_{jt}$ --- so, by construction, we assume such
correlations are constant across all asset pairs and over all
time.

The elements $C_{ij}$ ($i \neq j$) of the covariance matrix, $C
= \mtx{RR'}/T$, are found to be normally distributed with mean
\begin{eqnarray}
  \mu'_X &=& {\sigma^2 \over \sqrt{2\pi} (1 - \phi^2)(1 -
    \rho^2)^{1/4}} \left[ P^{-3/2}_{-1/2}(-\rho) -
    P^{-3/2}_{-1/2}(\rho)
  \right] \label{eq:gaussian_mean2}
\end{eqnarray}
and variance
\begin{eqnarray}
  \sigma'^2_X &=& {1 \over (1 - \phi^2)^2}\left[
    \sum_{t=1}^T \sum_{k=1}^{t-1} 2\left(
      \phi^k \over T
    \right)^2 \sigma^6 + \sum_{t=1}^T
    {\sigma^4 (1 - \rho^2)^2 \over T^2} v^2(\rho)
  \right] \nonumber \\
  &=& {2 \sigma^6 \over T (1 - \phi^2)^2} \left[
    {\phi^2 \over 1 - \phi^2} -
    {\phi^2 (1 - \phi^{2T}) \over
      T(1 - \phi^2)}
  \right] + {\sigma^4 (1 - \rho^2)^2 v^2(\rho) \over
    T (1 - \phi^2)^2} \nonumber \\
  &\approx& {2 \sigma^6 \phi^2 \over T (1 - \phi^2)^3}
  + {\sigma^4 (1 - \rho^2)^2 v^2(\rho) \over
    T (1 - \phi^2)^2} \label{eq:gaussian_variance2}
\end{eqnarray}
where $P^\mu_\nu(\cdot)$ is Ferrer's function of the first
kind. See appendix \ref{chp:gaussian_elements_dist}. For the
definition of Ferrer's function, see equation
\ref{eq:Ferrers_1st}.

Equation \ref{eq:gaussian_mean2} tells that, if
two return series $i$ and $j$ are not correlated, i.e. $\rho = 0$,
auto-correlation in the returns does not introduce a bias into the
estimation of their covariance, i.e. $\mu'_X$, since the difference
between the two Ferrer's functions evaluate to 0 in equation
\ref{eq:gaussian_mean2}; if, however, the return series are indeed
correlated, auto-correlation in the returns rescales the
covariance through a multiplicative factor $1/(1 - \phi^2)$.

In addition, equation \ref{eq:gaussian_variance2} tells that
auto-correlation in the returns always makes the covariance
estimation more noisy --- auto-correlation not only rescales the
variance of the no-autocorrelation estimation by $1/(1 - \phi^2)^2$
but even adds an extra term ${2 \sigma^6 \phi^2 \over T (1 -
  \phi^2)^3}$.

For the diagonal elements of the covariance matrix $C$ we have
\begin{eqnarray}
  \E(C_{ii}) &=& {1 \over T}\left[
    \sum_{k=0}^{t-1} \phi^{2k} \sigma^2
  \right] \nonumber \\
  &=& {\sigma^2 \over (1 - \phi^2) T} \left[
    T - {\phi^2 (1 - \phi^{2T}) \over 1 - \phi^2}
  \right] \nonumber \\
  & \approx & {\sigma^2 \over 1 - \phi^2} \left[
    1 - {\phi^2 \over T (1 - \phi^2)}
  \right] \label{eq:gaussian_cii_mean2}
\end{eqnarray}
and
\begin{eqnarray}
  \var(C_{ii}) &=& \sum_{t=1}^T \left[
    \sum_{k=0}^{t-1} {\phi^{4k} \sigma^4 \over T^2} 2 +
    \sum_{k,l=0}^{t-1} {\phi^{2(k+l)} \over T^2} \sigma^6
  \right] \nonumber \\
  &=& \sum_{t=1}^T \left[
    {2 \sigma^4 \over T^2} {1 - \phi^{4t} \over 1 - \phi^4} +
    {\sigma^6 \over T^2} \left(
      {1 - \phi^{2t} \over 1- \phi^2}
    \right)^2 \right] \nonumber \\
  &=& {2 \sigma^4 \over T (1 - \phi^4)} -
  {2 \sigma^4 \phi^4 (1 - \phi^{4T}) \over T^2(1 - \phi^4)^2} +
  \nonumber \\
  && {\sigma^6 \over T (1 -\phi^2)^2} -
  {2 \sigma^6 \phi^2 (1 - \phi^{2T}) \over T^2 (1 - \phi^2)^3} +
  {\sigma^6 \phi^4 (1 - \phi^{4T}) \over T^2 (1 - \phi^2)^2 (1 -
    \phi^4)} \nonumber \\
  &\approx& {2 \sigma^4 \over T (1 - \phi^4)} + {\sigma^6 \over T (1
    -\phi^2)^2} \label{eq:gaussian_cii_variance2}
\end{eqnarray}

From equation \ref{eq:gaussian_cii_mean2} we see that auto-correlation
in the returns increases the variance of the return series; and from
equation \ref{eq:gaussian_cii_variance2} we see that the variance of
that variance estimation is also increased by
auto-correlations. Moreover, we note that $\var(C_{ii})$ scales with T
approximately as $1/T$, similar to the behavior of
$\var(C_{ij})$. This is to be compared with the case of \gls{garch}
returns discussed in chapter \ref{chp:CrossCorrelationFat}.

\section{Distribution of the Eigenvalues}
\label{sec:GCC-numerical}
For a Wishart matrix $\mtx{RR'}$, theoretical results are available
for the eigenvalue distribution, and we summarize them in appendix
\ref{sec:wishart_eigen_dist}. In short, the joint probability density
function of the eigenvalues is given by equation
\ref{eq:wishart_eigen_pdf} when neither auto-correlation nor
cross-correlation is present in $\mtx R$. Moreover, the mariginal
distribution of the largest eigenvalue is approximated by a gamma
distribution \cite{Chiani2012}.

However, as detailed in the derivation leading to equation
\ref{eq:cross-corr-matrix-PDF}, the distribution of $\mtx{RR'}$ is not
Wishart when the columns of $\mtx{R}$ are correlated
(auto-correlation). Deriving the eigenvalue distribution analytically
in this case is beyond the scope of this thesis. Instead, we resort to
numerical methods.

As before we consider the \gls{ar}(1) process:
\begin{equation*}
  \vec{r}_t = \phi \vec{r}_{t-1} + \vec{a}_{t}
\end{equation*}
where $\vec{a}_t \sim N(0, \mtx{I})$, i.e. the elements of $\vec{a}_t$ are
independent Gaussian random variable with zero mean and unit
variance. Now we investigate how the eigenvalue distribution depends on
the auto-correlation strength parameter $\phi$. Figure
\ref{fig:GaussianMarkovSpectrumPDF} shows the results of the
simulation.
\begin{figure}[htb!]
  \begin{center}
    \includegraphics[scale=0.4, clip=true, trim=90 228 115
    226]{../pics/GaussianMarkovSpectrumPDF.pdf}
  \end{center}
  \vspace{-10mm}
  \caption{\small \it
      Eigenvalue distribution with correlation time $\tau$ ranging
      from 0 to 3. The 1st blue line, which is shown as stairs, is the
      theoretical eigenvalue distribution according to the
      Marcenko-Pastur law (see eq.\ref{eq:MP_pdf}). In the simulation
      we have chosen $q = N/T = 50/1000 = 0.05$ and the standard
      deviation of the returns $\sigma=1$. For each value of $\tau$ we
      generate 2000 instances of $N \times T$ random matrix $R$, and
      compute C as $C=RR'/T$. Hence each curve in the figure is
      constructed from 2000 sets of eigenvalues. The correspondance
      between the correlation time $\tau$ and the auto-regressive
      coefficient $\phi$ is $\tau = -\ln 2 / \ln \phi$.
    }
  \label{fig:GaussianMarkovSpectrumPDF}
\end{figure}
It is clear from the figure that the maximum eigenvalue moves
consistently to the right as the value of $\phi$ increases, and, as
shown in figure \ref{fig:Gaussian_mineig}, the minimum eigenvalue
also increases with $\phi$.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.4, clip=true, trim=100 226 116
  133]{../pics/Gaussian_mineig.pdf}
  \caption{\small \it The minimum eigenvalue versus
    auto-correlation strength $\phi$. For each value of $\phi$ 2000
    random matrices are generated and their eigenvalues are
    calculated. The minimum eigenvalue of each random matrix is noted
    and the mean of the 2000 such minimum eigenvalues are plotted
    against the chosen value of $\phi$. 20 values of $\phi$ are
    included in the plot, ranging from 0 to 0.95 with step size 0.05.}
  \label{fig:Gaussian_mineig}
\end{figure}

Chiani showed that the marginal distribution of the maximum eigenvalue
($\lambda_1$) is approximately gamma when neither cross-correlation
nor auto-correlation is present\cite{Chiani2012}. So we compare
in figure \ref{fig:GaussianMarkov005MaxEigCDF_loglog} the empirical
cumulative distribution function (CDF) of the maximum eigenvalue 
with the CDF of a gamma distribution. The cases where
auto-correlations are present ($\phi > 0$) have also been included.
\begin{figure}
  \begin{center}
    \includegraphics[scale=0.5, clip=true, trim=0 238 0
    197]{../pics/GaussianMarkov005MaxEigCDF_loglog.pdf}
  \end{center}
  \caption{\small \it Cummulative distribution function (CDF) of the
    maximum eigenvalue ($\lambda_1$) from numerical simulations (blue
    lines) are compared to the fitted gamma distribution (red
    lines). Each pair of CDFs correspond to a fixed autocorrelation
    strength ($\phi$). The parameters $k$ and $\theta$ of the gamma
    distribution are fit to data by matching the 2nd and the 3rd
    moments of the gamma distribution to the corresponding moments of
    the empirical distribution. Then the parameter $\alpha$ in
    equation \ref{eq:TracyWidom-Gamma} is chosen to be $\alpha =
    k\theta - \E\left({\lambda_1 - \mu_{NT} \over
        \sigma_{NT}}\right)$. The curves are plotted on log-log
    scale.}
  \label{fig:GaussianMarkov005MaxEigCDF_loglog}
  %\vspace{-10mm}
\end{figure}
It is seen in the figure that gamma distributions with different
parameters fit fairly well in all cases. So we conclude that a
gamma distribution not only approximates the maximum eigenvalue
distribution at the absence of autocorrelations but does so even at
the {\it presence} of autocorrelations.

Since the maximum eigenvalue distribution is approximated by a gamma
distribution characterized by parameters $k$, $\theta$, and $\alpha$
\footnotemark, the influence of the autocorrelations can be
characterized by the dependence of $k$, $\theta$, and  $\alpha$ on
$\phi$. While these dependences are rather intricate, good support can
be found in the data for the following approximate relation:
\footnotetext{
  The mean, variance and skewness of the gamma distribution are given
  by
  \begin{eqnarray*}
    \text{mean} &=& k\theta \\
    \text{variance} &=& k\theta^2 \\
    \text{skewness} &=& 2/\sqrt{k}
  \end{eqnarray*}
}
\begin{eqnarray}
  k\theta &=& a \tan^2{\pi \phi \over 2} + b\tan{\pi \phi \over 2} +
  c \label{eq:k_theta-phi}
\end{eqnarray}
Here we note that $k\theta$ is the mean of the gamma distribution.
To verify this relation, we first fit a 2nd order polynomial and
obtain the coefficients $a$, $b$, $c$; then for each data point
$k_n\theta_n$ we solve the quadratic equation
\begin{eqnarray}
  a \tan^2{\pi \phi'_n \over 2} + b\tan{\pi \phi'_n \over 2} + c -
  k_n\theta_n &=& 0\label{eq:k_theta-phi_2}
\end{eqnarray}
for $\tan{\pi \phi'_n \over 2}$. If relation \ref{eq:k_theta-phi} is a
good approximation, a close match between $\tan{\pi \phi'_n \over 2}$
and $\tan{\pi \phi_n \over 2}$ is expected. From figure 
\ref{fig:phi_and_roots} one can see this is indeed the case.

\begin{figure}[htb!]
  \centering
    \includegraphics[scale=0.5, clip=true, trim=37 217 39
    170]{../pics/phi_and_roots.pdf}
  \caption{\small \it Upper plot: $\tan{\pi \phi'_n \over 2}$ against
    $\tan{\pi \phi_n \over 2}$. The fitted line has equation $y_n =
    0.995 \tan{\pi \phi_n \over 2} + 0.0146$. Lower plot: Residuals of
    the linear fit, i.e. $\tan{\pi \phi'_n \over 2} - y_n$. 20 values
    of $\phi$ are included in the plot, ranging from 0 to 0.95 with
    step size 0.05.}
  \label{fig:phi_and_roots}
\end{figure}
